---
title: "TTC Delays"
author: "Ashish Adhikari"
date: "2025-05-11"
output:
  word_document: default
  pdf_document: default
---

# Exploratory Data Analysis (EDA)

## What are the most frequent delay causes (Code)?

## Which stations have the most delays?

## Are delays worse at certain times of day or days of week?


# Predictive Modeling

## Regression: Predict delay duration based on time, line, cause

## Classification: Classify whether a delay will be > 5 mins

<br>

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width = 7, fig.height = 5,
                      fig.path = 'Figs/', echo = TRUE)

#This sets the working directory
knitr::opts_knit$set(root.dir = 'C:/Users/ROG/Desktop/Canada/Projects/R/TTC Delay Data')
```


```{r, include=FALSE}

if(!is.null(dev.list()))dev.off()
rm(list = ls())
cat("\014")

options(scipen=9)

```


```{r, include=FALSE}
#Load packages

if(!require(tinytex)){install.packages("tinytex")}
library("tinytex")

if(!require(pastecs)){install.packages("pastecs")}
library("pastecs")

if(!require(lattice)){install.packages("lattice")}
library("lattice")

if(!require(vcd)){install.packages("vcd")}
library("vcd")

if(!require(HSAUR)){install.packages("HSAUR")}
library("HSAUR")

if(!require(rmarkdown)){install.packages("rmarkdown")}
library("rmarkdown")

if(!require(ggplot2)){install.packages("ggplot2")}
library("ggplot2")

if(!require(polycor)){install.packages("polycor")}
library("polycor")

if(!require(nnet)){install.packages("nnet")}
library("nnet")

if(!require(klaR)){install.packages("klaR")}
library("klaR")

if(!require(partykit)){install.packages("partykit")}
library("partykit")

```

# 1. Preliminary and Exploratory


```{r}
# Load data 
df <- read.table("ttc-subway-delay-data-2024.csv",
                 header=TRUE,
                 sep=",",
                 quote = "\"",
                 fill = TRUE,
                 stringsAsFactors = FALSE)
df <- as.data.frame(df)

```
<br>

## 2 Data Description

<br>
```{r}
head(df)

```
<br>

```{r}
summary(df)
```

```{r}
df <- na.omit(df)

```

```{r}
summary(df)
```


We know from looking in the datasheet, that **Date** is the date of records in *YYYY-MM-DD** format, and **Time** is the time of the incident in **hrs** format.

```{r}
# Change Date to MM/DD/YYYY
df$Date <- as.Date(df$Date)

```

Since we have time in hrs, we will break it down to different time ranges as:
* Early Morning: 5 to 9
* Morning: 9 to 12
* Afternoon: 12 to 17
* Evening: 17 to 21
* Night: 21 to 5


```{r}
# Extract Hour and convert them to numeric
df$Hour <- as.numeric(sub(":.*", "", df$Time))

# Assign Hours to TimeRange as defined above
df$TimeRange <- with(df, ifelse(Hour >= 5 & Hour < 9, "Early Morning",
                         ifelse(Hour >= 9 & Hour < 12, "Morning",
                         ifelse(Hour >= 12 & Hour < 17, "Afternoon",
                         ifelse(Hour >= 17 & Hour < 21, "Evening", "Night")))))

```


```{r}
summary(df)
```
We can see that we have two new columns *Hour* and *TimeRange*, with Hour being numeric min value 0 and max value 23 and TimeRange being character. Now we don't need Time and Hour columns so we will drop them.

We know from the separate Code Description file that 
* *Day* is the name of the day of the week
* *Code* can be one of 129 or 71 different delay codes 
* *Bound* is the direction of the train on the line
* *Line* is the TTC Subway Line
* *Vehicle* is the TTC train number, and has no numeric value

So we convert them all including TimeRange to factors

```{r}
df <- df[ ,-2]
df <- df[ ,-10]

df$Day <- as.factor(df$Day)
df$Station <- as.factor(df$Station)
df$Code <- as.factor(df$Code)
df$Bound <- as.factor(df$Bound)
df$Line <- as.factor(df$Line)
df$Vehicle <- as.factor(df$Vehicle)
df$TimeRange <- as.factor(df$TimeRange)
```

```{r}
summary(df)
```

```{r}
levels(df$Line)
```
There are some inconsistencies in the **Line** column, so we will standardize them. And it also has "None" value in it which cannot be the case, we will treat them as missing values and remove them. 

YUS stands for Yonge–University–Spadina Line, which was the older name for what is now officially called the Yonge–University Line, abbreviated as YU.

```{r}
df$Line <- as.character(df$Line)

# Replace known variations
df$Line[df$Line %in% c("YUS", "LINE 1", "ONGE-UNIVERSITY AND BL")] <- "YU"
df$Line[df$Line %in% c("BLOOR DANFORTH")] <- "BD"
df$Line[df$Line %in% c("SHEP")] <- "SHP"

# Replace all combo labels with "YU-BD"
df$Line[df$Line %in% c("YU / BD", "YU/ BD", "YU/BD", "YU & BD", "YUS/BD", "BD/YU", "BD / YU", "BD/ YU", "BD/ YUS", "YU/BD/SHP")] <- "YU-BD"

# Remove irrelevant labels
df$Line[df$Line %in% c("TRACK LEVEL ACTIVITY", "109 RANEE", "20 CLIFFSIDE", "")] <- NA
df <- df[!is.na(df$Line), ]

# Convert back to factor
df$Line <- factor(df$Line)

# Check final cleaned levels
levels(df$Line)

```

```{r}
summary(df)
```

```{r}
levels(df$Code)

sum(df$Code %in% c("XXXXX"))
```
Now looking into Code we found one entry with invalid value, so we simply remove them. 
```{r}
df$Code <- as.character(df$Code)

# Remove invalid codes
df$Code[df$Code %in% c("XXXXX")] <- NA
df <- df[!is.na(df$Code), ]

df$Code <- factor(df$Code)
levels(df$Code)

```
```{r}
summary(df)
```
Now in *Bound*, there is "B" which is ambiguous and missing values. Similarly, in case of Vehicle as well "0" is definitely missing values. 

Significant number of rows are missing in both Bound and Vehicle, so we cannot simply drop them. We keep values with None for now. But B is just 29 observations so we drop all observations with B.

```{r}
df$Bound <- as.character(df$Bound)

# Drop rows where Bound == "B"
df <- df[df$Bound != "B", ]

df$Bound <- as.factor(df$Bound)
```


```{r}
summary(df)
```




```{r}
# Prepare data in long format
delay_data <- data.frame(
  Category = c(rep("Min.Delay", length(df$Min.Delay)),
               rep("Min.Gap", length(df$Min.Gap))),
  Value = c(df$Min.Delay, df$Min.Gap)
)

# Plot
ggplot(delay_data, aes(x = Category, y = Value)) +
  geom_boxplot(fill = "lightgreen") +
  labs(title = "Boxplot of Min.Delay and Min.Gap",
       x = "",
       y = "Minutes") +
  theme_minimal()

```

There are few delays and gaps with time greater than 50 minutes, they might be rare and standalone incidents so we simply remove them.


```{r}
df <- df[df$Min.Delay <= 50 & df$Min.Gap <= 50, ]

```

```{r}
# Prepare data in long format
delay_data <- data.frame(
  Category = c(rep("Min.Delay", length(df$Min.Delay)),
               rep("Min.Gap", length(df$Min.Gap))),
  Value = c(df$Min.Delay, df$Min.Gap)
)

# Plot
ggplot(delay_data, aes(x = Category, y = Value)) +
  geom_boxplot(fill = "lightgreen") +
  labs(title = "Boxplot of Min.Delay and Min.Gap",
       x = "",
       y = "Minutes") +
  theme_minimal()

```


```{r}
stat.desc(df[, c("Min.Delay", "Min.Gap")])

```

```{r}
par(mfrow = c(2, 2))

hist(df$Min.Delay, 
   main = "Delay (in Minutes)", 
   xlab = "Value", 
   ylab = "Frequency", 
   col = "#ADD8E6", 
   freq = FALSE) # `freq = FALSE` scales the histogram for overlaying density
lines(density(df$Min.Delay))

hist(df$Min.Gap, 
   main = "Gap (in Minutes)", 
   xlab = "Value", 
   ylab = "Frequency", 
   col = "#ADD8E6", 
   freq = FALSE) # `freq = FALSE` scales the histogram for overlaying density
lines(density(df$Min.Gap))
```

Since there are a lot of reported incidents that have caused 0 mins delay, it might be a good idea to remove them from our data frame as it skews our data and can create bias in our models later on.

```{r}
df <- df[df$Min.Delay > 0 & df$Min.Gap > 0, ]

```

Checking for correlation: 

```{r}
cr <- cor(df[sapply(df, is.numeric)], method="spearman")
round(cr, 2)

```
Since Delay of one train and the Gap between trains are highly correlated, we can drop one. Here we chose to drop Min.Gap.

```{r}
df <- df[, -6]
summary(df)
```



## What are the most frequent incidents (Code)?

```{r}
code_freq <- table(df$Code)

top5 <- sort(code_freq, decreasing = TRUE)[1:5]

top5_df <- data.frame(Code = names(top5), Frequency = as.numeric(top5))

ggplot(top5_df, aes(x = reorder(Code, Frequency), y = Frequency)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 5 Delay Causes",
       x = "Delay Code",
       y = "Number of Delays") +
  theme_minimal()


```
```{r}

code_legend <- read.csv("Code Descriptions.csv", stringsAsFactors = FALSE)
top5_description <- merge(top5_df, code_legend, by = "Code")
top5_description

```

The most common causes of delays are disorderly patron, injured/ill customer in station or some other security concern.


## Which stations have the most delays?

```{r}
station_counts <- sort(table(df$Station), decreasing = TRUE)
top_stations <- data.frame(Station = names(station_counts), Frequency = as.numeric(station_counts))

ggplot(head(top_stations, 10), aes(x = reorder(Station, Frequency), y = Frequency)) +
  geom_bar(stat = "identity", fill = "tomato") +
  coord_flip() +
  labs(title = "Top 10 Delayed Stations",
       x = "Station",
       y = "Number of Delays") +
  theme_minimal()

```



```{r}
par(mfrow = c(2, 2))

hist(df$Min.Delay, 
   main = "Delay (in Minutes)", 
   xlab = "Value", 
   ylab = "Frequency", 
   col = "#ADD8E6", 
   freq = FALSE) # `freq = FALSE` scales the histogram for overlaying density
lines(density(df$Min.Delay))

```


# Regression: Predict delay duration 

```{r}
# 70% train and 30 % test
sr<- 0.70

# Find number of rows of data
n.row <- nrow(df)
n.row
```
#### Choosing rows for the training sample
  
```{r}

set.seed(8848) 
training.rows <- sample(1:n.row, sr*n.row, replace=FALSE)

#Assign to the training sample

train <- subset(df[training.rows,])

#Assign the balance to the test sample

test <- subset(df[-c(training.rows),])

summary(train)
summary(test)
```

## Linear Regression

### Model using all variables
  
  
```{r, include = FALSE}
mod.All <- lm(Min.Delay ~ ., data=train)
summary(mod.All)
```

```{r}

pred <- predict(mod.All, newdata=train)

RMSE_trn_All <- sqrt(mean((train$Min.Delay - pred)^2))
round(RMSE_trn_All, 2)

pred <- predict(mod.All, newdata=test)

RMSE_tst_All <- sqrt(mean((test$Min.Delay - pred)^2))
round(RMSE_tst_All, 2)

round(RMSE_trn_All - RMSE_tst_All, 2)
```
For model using **all variables**, 
  
We can see that the p-value of **F-Test** is less than 0.05, so we determine that the model **has some value.**
  
The **Adjusted R^2** shows that the model explains just **23.6%** of variation in the dependent variable.
  
Looking at the **Residuals** looking at its symmetry we'll say that the mean is **around 0**, but the distribution is not symmetrical. 
  
From **T-test** for the intercept and all the variables, we have p-value for all except few Stations and Codes. 
  
Now, looking at the coefficients for **sensibility**, 
  
The **RMSE** for train and test data seems consistent. 



## Model using backward regression
  
```{r, include = FALSE}
mod.Back = step(mod.All, direction="backward", details=TRUE)
summary(mod.Back)
```

```{r}
pred <- predict(mod.Back, newdata=train)

RMSE_trn_Back <- sqrt(mean((train$Min.Delay - pred)^2))
round(RMSE_trn_Back, 2)

pred <- predict(mod.Back, newdata=test)

RMSE_tst_Back <- sqrt(mean((test$Min.Delay - pred)^2))
round(RMSE_tst_Back, 2)

round(RMSE_trn_Back - RMSE_tst_Back, 2)
```


```{r}
par(mfrow = c(2, 2)) 
plot(mod.Back, col=3, main="Full Model")  
par(mfrow = c(1, 1)) 
```


```{r}

# Scale numeric inputs (needed for neural nets)
train_scaled <- train
test_scaled <- test

# Convert target to scaled version
max_delay <- max(train$Min.Delay)
train_scaled$Min.Delay <- train$Min.Delay / max_delay
test_scaled$Min.Delay  <- test$Min.Delay / max_delay

# Fit neural net (simplified)
nn_model <- nnet(Min.Delay ~ TimeRange + Day + Line + Code,
                 data = train_scaled, size = 7, linout = TRUE)

# Predict and rescale
pred_nn <- predict(nn_model, newdata = train_scaled) * max_delay

# RMSE
rmse_nn <- sqrt(mean((pred_nn - train$Min.Delay)^2))
rmse_nn

# Predict and rescale
pred_nn <- predict(nn_model, newdata = test_scaled) * max_delay

# RMSE
rmse_nn <- sqrt(mean((pred_nn - test$Min.Delay)^2))
rmse_nn

```


