---
title: "TTC Delays"
author: "Ashish Adhikari"
date: "2025-05-11"
output:
  word_document: default
  pdf_document: default
---

# Exploratory Data Analysis (EDA)

## What are the most frequent delay causes (Code)?

## Which stations have the most delays?

## Are delays worse at certain times of day or days of week?

## What are the average and max delay durations?

## Delay frequency by hour vs. weekday


# Predictive Modeling

## Regression: Predict delay duration based on time, line, cause

## Classification: Classify whether a delay will be > 5 mins


```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width = 7, fig.height = 5,
                      fig.path = 'Figs/', echo = TRUE)

#This sets the working directory
knitr::opts_knit$set(root.dir = 'C:/Users/ROG/Desktop/Canada/Projects/R/TTC Delay Data')
```


```{r, include=FALSE}

if(!is.null(dev.list()))dev.off()
rm(list = ls())
cat("\014")

options(scipen=9)

```


```{r, include=FALSE}
#Load packages

if(!require(tinytex)){install.packages("tinytex")}
library("tinytex")

if(!require(pastecs)){install.packages("pastecs")}
library("pastecs")

if(!require(lattice)){install.packages("lattice")}
library("lattice")

if(!require(vcd)){install.packages("vcd")}
library("vcd")

if(!require(HSAUR)){install.packages("HSAUR")}
library("HSAUR")

if(!require(rmarkdown)){install.packages("rmarkdown")}
library("rmarkdown")

if(!require(ggplot2)){install.packages("ggplot2")}
library("ggplot2")

if(!require(polycor)){install.packages("polycor")}
library("polycor")

if(!require(nnet)){install.packages("nnet")}
library("nnet")

if(!require(klaR)){install.packages("klaR")}
library("klaR")

if(!require(partykit)){install.packages("partykit")}
library("partykit")

```

# 1. Preliminary and Exploratory


```{r}
# Load data 
df <- read.table("TTC Subway Delay Data since 2025.csv",
                 header=TRUE,
                 sep=",",
                 quote = "\"",
                 fill = TRUE,
                 stringsAsFactors = FALSE)
df <- as.data.frame(df)

```
<br>

## 2 Data Description

<br>
```{r}
head(df)

summary(df)
```

We see that **X=id** is just list of numbers for the rows, min value 1 to max value 6357 which is the number of observations in our datasheet. So we can drop this column as it doesn't give us any value in our operations. 

Everything else is given as character. But we know from looking in the datasheet, that **Date** is the date of records in *YYYY-MM-DD** format, and **Time** is the time of the incident in **hrs** format.

```{r}
# Droping the X_id
df <- df[, -1]

# Change Date to MM/DD/YYYY
df$Date <- as.Date(df$Date)

```

Since we have time in hrs, we will break it down to different time ranges as:
* Early Morning: 5 to 9
* Morning: 9 to 12
* Afternoon: 12 to 17
* Evening: 17 to 21
* Night: 21 to 5


```{r}
# Extract Hour and convert them to numeric
df$Hour <- as.numeric(sub(":.*", "", df$Time))

# Assign Hours to TimeRange as defined above
df$TimeRange <- with(df, ifelse(Hour >= 5 & Hour < 9, "Early Morning",
                         ifelse(Hour >= 9 & Hour < 12, "Morning",
                         ifelse(Hour >= 12 & Hour < 17, "Afternoon",
                         ifelse(Hour >= 17 & Hour < 21, "Evening", "Night")))))

```


```{r}
summary(df)
```
We can see that we have two new columns *Hour* and *TimeRange*, with Hour being numeric min value 0 and max value 23 and TimeRange being character. Now we don't need Time and Hour columns so we will drop them.

We know from the separate Code Description file that 
* *Day* is the name of the day of the week
* *Code* can be one of 140 different delay codes 
* *Bound* is the direction of the train on the line
* *Line* is the TTC Subway Line
* *Vehicle* is the TTC train number, and has no numeric value

So we convert them all including TimeRange to factors

```{r}
df <- df[ ,-2]
df <- df[ ,-10]

df$Day <- as.factor(df$Day)
df$Station <- as.factor(df$Station)
df$Code <- as.factor(df$Code)
df$Bound <- as.factor(df$Bound)
df$Line <- as.factor(df$Line)
df$Vehicle <- as.factor(df$Vehicle)
df$TimeRange <- as.factor(df$TimeRange)
```

```{r}
summary(df)
```

```{r}
levels(df$Line)
```
There are some inconsistencies in the **Line** column, so we will standardize them. And it also has "None" value in it which cannot be the case, we will treat them as missing values and remove them. 

YUS stands for Yonge–University–Spadina Line, which was the older name for what is now officially called the Yonge–University Line, abbreviated as YU.

```{r}
df$Line <- as.character(df$Line)

# Removing "None" values from Line
df <- df[df$Line != "None", ]

# Replacing known variations with standard labels
df$Line[df$Line == "YUS"] <- "YU"
df$Line[df$Line %in% c("YU / BD", "YU/ BD", "YU/BD", "YUS/ BD", "YUS/BD")] <- "YU-BD"

df$Line <- factor(df$Line)

levels(df$Line)

```

```{r}
summary(df)
```

Now in *Bound*, there is B which is ambiguous and None which is definitely missing values. Similarly, in case of Vehicle as well None is definitely missing values. 

Significant number of rows are missing in both Bound and Vehicle, so we cannot simply drop them. We keep values with None for now. But B is just 4 observations so we drop all observations with B.

```{r}
df$Bound <- as.character(df$Bound)
df$Vehicle <- as.character(df$Vehicle)

# Drop rows where Bound == "B"
df <- df[df$Bound != "B", ]

df$Bound <- as.factor(df$Bound)
df$Vehicle <- as.factor(df$Vehicle)
```


```{r}
summary(df)
```
We know Delay and Gap are in minutes, so we convert them to numeric. There None in those columns as well so we convert them to NA. 

```{r}
# Replace "None" with NA
df$Min.Delay[df$Min.Delay == "None"] <- NA
df$Min.Gap[df$Min.Gap == "None"]     <- NA

# Convert to numeric
df$Min.Delay <- as.numeric(df$Min.Delay)
df$Min.Gap   <- as.numeric(df$Min.Gap)

```

```{r}
summary(df)

```

## What are the most frequent delay causes (Code)?

```{r}
code_freq <- table(df$Code)

top5 <- sort(code_freq, decreasing = TRUE)[1:5]

top5_df <- data.frame(Code = names(top5), Frequency = as.numeric(top5))

ggplot(top5_df, aes(x = reorder(Code, Frequency), y = Frequency)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 5 Delay Causes",
       x = "Delay Code",
       y = "Number of Delays") +
  theme_minimal()


```
```{r}

code_legend <- read.csv("Code Descriptions.csv", stringsAsFactors = FALSE)
top5_description <- merge(top5_df, code_legend, by = "Code")
top5_description

```


The most common causes of delays are disorderly patron, injured/ill customer in station or some other security concern.


## Which stations have the most delays?

```{r}
station_counts <- sort(table(df$Station), decreasing = TRUE)
top_stations <- data.frame(Station = names(station_counts), Frequency = as.numeric(station_counts))

ggplot(head(top_stations, 10), aes(x = reorder(Station, Frequency), y = Frequency)) +
  geom_bar(stat = "identity", fill = "tomato") +
  coord_flip() +
  labs(title = "Top 10 Delayed Stations",
       x = "Station",
       y = "Number of Delays") +
  theme_minimal()

```


## Are delays worse at certain times of day or days of week?
```{r}
df <- df[df$Min.Delay <= 250 & df$Min.Gap <= 250, ]

```


```{r}
# Prepare data in long format
delay_data <- data.frame(
  Category = c(rep("Min.Delay", length(df$Min.Delay)),
               rep("Min.Gap", length(df$Min.Gap))),
  Value = c(df$Min.Delay, df$Min.Gap)
)

# Plot
ggplot(delay_data, aes(x = Category, y = Value)) +
  geom_boxplot(fill = "lightgreen") +
  labs(title = "Boxplot of Min.Delay and Min.Gap",
       x = "",
       y = "Minutes") +
  theme_minimal()

```



```{r}
# df <- na.omit(df)

```

