---
title: "TTC Delays"
author: "Ashish Adhikari"
date: "2025-05-11"
output:
  word_document: default
  pdf_document: default
---

# Exploratory Data Analysis (EDA)

## What are the most frequent delay causes (Code)?

## Which stations have the most delays?


# Predictive Modeling

## Classification: Classify whether a delay will be > 5 mins

<br>

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width = 7, fig.height = 5,
                      fig.path = 'Figs/', echo = TRUE)

#This sets the working directory
knitr::opts_knit$set(root.dir = 'C:/Users/ROG/Desktop/Canada/Projects/R/TTC Delay Data/ttc-delay-data')
```


```{r, include=FALSE}

if(!is.null(dev.list()))dev.off()
rm(list = ls())
cat("\014")

options(scipen=9)

```


```{r, include=FALSE}
#Load packages

if(!require(tinytex)){install.packages("tinytex")}
library("tinytex")

if(!require(pastecs)){install.packages("pastecs")}
library("pastecs")

if(!require(lattice)){install.packages("lattice")}
library("lattice")

if(!require(vcd)){install.packages("vcd")}
library("vcd")

if(!require(HSAUR)){install.packages("HSAUR")}
library("HSAUR")

if(!require(rmarkdown)){install.packages("rmarkdown")}
library("rmarkdown")

if(!require(ggplot2)){install.packages("ggplot2")}
library("ggplot2")

if(!require(polycor)){install.packages("polycor")}
library("polycor")

if(!require(nnet)){install.packages("nnet")}
library("nnet")

if(!require(klaR)){install.packages("klaR")}
library("klaR")

if(!require(partykit)){install.packages("partykit")}
library("partykit")

```

# 1. Preliminary and Exploratory


```{r}
# Load data 
df <- read.table("ttc-subway-delay-data-2024.csv",
                 header=TRUE,
                 sep=",",
                 quote = "\"",
                 fill = TRUE,
                 stringsAsFactors = FALSE)
df <- as.data.frame(df)

```
<br>

## 2 Data Description

<br>
```{r}
head(df)

```
<br>

```{r}
summary(df)
```

```{r}
df <- na.omit(df)

```

```{r}
summary(df)
```


We know from looking in the datasheet, that **Date** is the date of records in *YYYY-MM-DD** format, and **Time** is the time of the incident in **hrs** format.

```{r}
# Change Date to MM/DD/YYYY
df$Date <- as.Date(df$Date)

```

Since we have time in hrs, we will break it down to different time ranges as:
* Early Morning: 5 to 9
* Morning: 9 to 12
* Afternoon: 12 to 17
* Evening: 17 to 21
* Night: 21 to 5


```{r}
# Extract Hour and convert them to numeric
df$Hour <- as.numeric(sub(":.*", "", df$Time))

# Assign Hours to TimeRange as defined above
df$TimeRange <- with(df, ifelse(Hour >= 5 & Hour < 9, "Early Morning",
                         ifelse(Hour >= 9 & Hour < 12, "Morning",
                         ifelse(Hour >= 12 & Hour < 17, "Afternoon",
                         ifelse(Hour >= 17 & Hour < 21, "Evening", "Night")))))

```


```{r}
summary(df)
```
We can see that we have two new columns *Hour* and *TimeRange*, with Hour being numeric min value 0 and max value 23 and TimeRange being character. Now we don't need Time and Hour columns so we will drop them.
<br>
 
We know from the separate Code Description file that <br>
* *Day* is the name of the day of the week <br> 
* *Code* can be one of 129 or 71 different delay codes <br>
* *Bound* is the direction of the train on the line <br> 
* *Line* is the TTC Subway Line <br> 
* *Vehicle* is the TTC train number, and has no numeric value <br> 
<br> 
So we convert them all including TimeRange to factors

```{r}
df <- df[ ,-2]
df <- df[ ,-10]

df$Day <- as.factor(df$Day)
df$Station <- as.factor(df$Station)
df$Code <- as.factor(df$Code)
df$Bound <- as.factor(df$Bound)
df$Line <- as.factor(df$Line)
df$Vehicle <- as.factor(df$Vehicle)
df$TimeRange <- as.factor(df$TimeRange)
```

```{r}
summary(df)
```

```{r}
levels(df$Line)
```
There are some inconsistencies in the **Line** column, so we will standardize them. And it also has "None" value in it which cannot be the case, we will treat them as missing values and remove them. 

YUS stands for Yonge–University–Spadina Line, which was the older name for what is now officially called the Yonge–University Line, abbreviated as YU.

```{r}
df$Line <- as.character(df$Line)

# Replace known variations
df$Line[df$Line %in% c("YUS", "LINE 1", "ONGE-UNIVERSITY AND BL")] <- "YU"
df$Line[df$Line %in% c("BLOOR DANFORTH")] <- "BD"
df$Line[df$Line %in% c("SHEP")] <- "SHP"

# Replace all combo labels with "YU-BD"
df$Line[df$Line %in% c("YU / BD", "YU/ BD", "YU/BD", "YU & BD", "YUS/BD", "BD/YU", "BD / YU", "BD/ YU", "BD/ YUS", "YU/BD/SHP")] <- "YU-BD"

# Remove irrelevant labels
df$Line[df$Line %in% c("TRACK LEVEL ACTIVITY", "109 RANEE", "20 CLIFFSIDE", "")] <- NA
df <- df[!is.na(df$Line), ]

# Convert back to factor
df$Line <- factor(df$Line)

# Check final cleaned levels
levels(df$Line)

```

```{r}
summary(df)
```
  

SRT is the Scarborouh RT line that was decommissioned in 2023, so we can drop that and YU-BD is Interchane delays involving both Line 1 (YU) and Line 2 (BD). There aren't many observations there, we will keep them for now.
  
```{r}

# Remove rows where Line is "SRT"
df <- df[df$Line != "SRT", ]

# Drop unused factor levels
df$Line <- droplevels(df$Line)

# Check final levels
levels(df$Line)
```

```{r, include=FALSE}
levels(df$Code)
```


Now looking into Code we found one entry with invalid value, so we simply remove them. We also have,

 - EU* includes all delays caused by faulty train components, brakes, doors, air, voltage issues — grouped into Equipment Issue.
  
 - MU* includes injuries, no staff available, ESA-related, weather, etc. — best described as Operational/Medical.
  
 - PU* includes power failure, signals, controllers — so Signal/Power/Comms.
  
 - SU* includes assaults, police holds, robbery, suspicious package — clearly Security/Police.
  
 - TU* covers operator error, overspeed, control errors — best labeled as Operator Error / Transit Control.


```{r}
df$Code <- as.character(df$Code)

# Remove invalid codes
df$Code[df$Code %in% c("", "XXXXX")] <- NA

# Group by prefix-based categories aligned with descriptions
df$CodeGroup <- NA

# EU = Equipment-related (RC&S + mechanical)
df$CodeGroup[grepl("^EU", df$Code)] <- "Equipment Issue"

# MU = Miscellaneous Utility (includes injuries, ESA, etc.)
df$CodeGroup[grepl("^MU", df$Code)] <- "Operational/Medical"

# PU = Power, Communications, Signal-related
df$CodeGroup[grepl("^PU", df$Code)] <- "Signal/Power/Comms"

# SU = Security or Police-related
df$CodeGroup[grepl("^SU", df$Code)] <- "Security/Police"

# TU = Operator or Transit related errors
df$CodeGroup[grepl("^TU", df$Code)] <- "Operator Error / Transit Control"

# Everything else
df$CodeGroup[is.na(df$CodeGroup)] <- "Other"

# Drop any rows with missing Code
df <- df[!is.na(df$Code), ]

# Convert back to factors
df$Code <- factor(df$Code)
df$CodeGroup <- factor(df$CodeGroup)

```


```{r}
summary(df)
```
Now in *Bound*, there is "B" which is ambiguous and missing values. Similarly, in case of Vehicle as well "0" is definitely missing values. 
<br>
Significant number of rows are missing in both Bound and Vehicle, so we cannot simply drop them. We keep values with None for now. But B is just 29 observations so we drop all observations with B.
<br>
And we also drop the *code* column as we don't need them now.

```{r}
df$Bound <- as.character(df$Bound)

# Drop rows where Bound == "B"
df <- df[df$Bound != "B", ]

df$Bound <- as.factor(df$Bound)

# Droppin the code column

df <- df[ ,-4]
```


```{r}
summary(df)
```

<br>
There is a lot of inconsistent and dirty data in the *Station* Column. Since we do not have the domain expertise for standardizing station names, we will clean and standardize them using prompt from ChatGPT.

```{r}
# -------------------------------
# 1. Clean and simplify Station names
# -------------------------------

# Convert Station to uppercase and trim spaces
df$Station <- toupper(trimws(df$Station))

# Remove common suffixes like " STATION", text after dashes, and redundant line codes
df$Station <- gsub(" STATION.*", "", df$Station)
df$Station <- gsub(" BD$", "", df$Station)
df$Station <- gsub(" YU$", "", df$Station)
df$Station <- gsub(" -.*", "", df$Station)
df$Station <- gsub(" +", " ", df$Station)
df$Station <- trimws(df$Station)

# -------------------------------
# 2. Remove irrelevant/general stations or vague entries
# -------------------------------

irrelevant_patterns <- c(
  "LINE", "YARD", "WYE", "SHOP", "BUILDING", "CAR HOUSE", "CARHOUSE", "COMPLEX",
  "SUBWAY", "SUBSTATION", "CLOSURE", "SYSTEM", "UNKNOWN", "TRACK", "DIVISION", 
  "HOSTLER", "TRAINING", "EE", "EMERGENCY", "APPROACHING", "TO", "BETWEEN", 
  "AND", "VARIOUS", "CHANGE OVER", "OPERATIONS", "UNION TO", "YU-", "YU & BD"
)

# Remove rows that contain any of the above patterns
remove_rows <- grepl(paste(irrelevant_patterns, collapse = "|"), df$Station)
df <- df[!remove_rows, ]

# -------------------------------
# 3. Re-factor Station
# -------------------------------
df$Station <- factor(df$Station)

```

<br>

```{r}
levels(df$Station)
```
  
The station names are much cleaner now, but some duplicates, typos, and inconsistent naming remain.

```{r}
df$Station <- as.character(df$Station)

# Fix common typos and duplicates
df$Station[df$Station == "KENNDY"] <- "KENNEDY"
df$Station[df$Station == "KILPING"] <- "KIPLING"
df$Station[df$Station == "ST GEOGE"] <- "ST GEORGE"
df$Station[df$Station == "ST. CLAIR"] <- "ST CLAIR"
df$Station[df$Station == "ST. CLAIR WEST"] <- "ST CLAIR WEST"
df$Station[df$Station == "ST. GEORGE"] <- "ST GEORGE"
df$Station[df$Station == "WISLON"] <- "WILSON"
df$Station[df$Station == "YOUNGE-UNIVERSITY-SPAD"] <- "YONGE-UNIVERSITY-SPADINA"
df$Station[df$Station == "FINCH STN 39 BUS PLATF"] <- "FINCH"
df$Station[df$Station %in% c("DAVISVILLE BUILD-UP", "DAVISVILLE BUILD UP", "DAVISVILLE BUILDUP")] <- "DAVISVILLE"
df$Station[df$Station == "OLD MILLS"] <- "OLD MILL"
df$Station[df$Station %in% c("SHEPPARD YONGE", "SHEPPARD-YONGE")] <- "SHEPPARD-YONGE"
df$Station[df$Station == "SPADINA YUS"] <- "SPADINA"
df$Station[df$Station == "BLOOR SATION"] <- "BLOOR"
df$Station[df$Station == "DOWNVIEW PARK STN"] <- "DOWNSVIEW PARK"
df$Station[df$Station == "FINCH WEST CENTER TRAC"] <- "FINCH WEST"
df$Station[df$Station == "LAWRENCE WEST CENTER T"] <- "LAWRENCE WEST"
df$Station[df$Station == "OSGOODE POCKET"] <- "OSGOODE"
df$Station[df$Station %in% c("SCARBOROUGH CENTRE STA", "SCARBOROUGH CTR STATIO")] <- "SCARBOROUGH CENTRE"
df$Station[df$Station == "SHEBOURNE"] <- "SHERBOURNE"
df$Station[df$Station %in% c("ST GEORGE BD/YU STATIO", "ST GEORGE YUS")] <- "ST GEORGE"
df$Station[df$Station == "YORK UNIVERSITY STATIO"] <- "YORK UNIVERSITY"
df$Station[df$Station == "KING COMMERCE"] <- "KING"
df$Station[df$Station == "VAUGHAN MC"] <- "VMC"
df$Station[df$Station == "NORTH YORK CTR"] <- "NORTH YORK CENTRE"
df$Station[df$Station == "SHEPPARD TAIL"] <- NA  # if not a valid stop
df$Station[df$Station == "YONGE"] <- "YONGE-UNIVERSITY-SPADINA"

df <- df[!is.na(df$Station), ]
df$Station <- factor(df$Station)

# Collapse YONGE variants
df$Station[df$Station %in% c(
  "YONGE-UNIVERISTY & BLO", "YONGE-UNIVERSITY", "YONGE-UNIVERSITY-SPADI", 
  "YONGE-UNIVERSITY & BLO", "YONGE-UNIVERSITY SUBWA", "YONGE / UNIVERSITY LIN",
  "YONGE UNIVERSITY", "YONGE UNIVERSITY / BLO", "YONGE UNIVERSITY SUBWA",
  "YONGE/UNIVERSITY/SPADI"
)] <- "YONGE-UNIVERSITY-SPADINA"

# Drop ambiguous or irrelevant entries
drop_stations <- c(
  "JANE & RUNNYMEDE STATI",
  "KENNEDY, DON MILLS, SC",
  "LOST ARTICLES OFFICE",
  "MCDONALD CARTIER EMERG",
  "MULTIPLE"
)
df <- df[!df$Station %in% drop_stations, ]

# Remove non-public stations
df <- df[!df$Station %in% c(
  "111 SPADINA ROAD", 
  "1900 YONGE MCBRIEN BLD", 
  "1900 YONGE ST- MCBRIEN", 
  "2233 SHEPPARD WEST", 
  "ALL", 
  "BUCHAN ESB"
), ]

# merge "YONGE-UNIVERSITY-SPADINA" into "YONGE-UNIVERSITY" 
df$Station <- as.character(df$Station)
df$Station[df$Station == "YONGE-UNIVERSITY-SPADINA"] <- "YONGE-UNIVERSITY"
df$Station <- factor(df$Station)

# Refactor again
df$Station <- factor(df$Station)

```

```{r}
levels(df$Station)
```
The station looks clean and now we can perform station specific analysis.

Visualize stations with most delays:

```{r}
# Count frequency of each station
station_counts <- as.data.frame(table(df$Station))

# Order by descending frequency
station_counts <- station_counts[order(-station_counts$Freq), ]

# Take top 5
top5_stations <- head(station_counts, 5)

# Bar plot
library(ggplot2)
ggplot(top5_stations, aes(x = reorder(Var1, -Freq), y = Freq)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Top 5 Stations with Most Delays",
       x = "Station",
       y = "Number of Delays") +
  theme_minimal()


```

now we further create a WeekGroup variable for WeekDay and WeekEnd, then we can drop "Day" column

```{r}
# Create new WeekGroup variable
df$Day <- as.character(df$Day)

df$WeekGroup <- ifelse(df$Day %in% c("Saturday", "Sunday"), "Weekend", "Weekday")

df$WeekGroup <- factor(df$WeekGroup)
df <- df[, -2]

summary(df)
```



```{r}

# Gather Min.Delay and Min.Gap into one long-format data frame
delay_data <- data.frame(
  Category = rep(c("Min.Delay", "Min.Gap"), each = nrow(df)),
  Value = c(df$Min.Delay, df$Min.Gap)
)

# Box-Plot of Minutes of Delay and Gap
ggplot(delay_data, aes(x = Category, y = Value)) +
  geom_boxplot(fill = "lightgreen") +
  labs(title = "Boxplot of Min.Delay and Min.Gap",
       x = "",
       y = "Minutes") +
  theme_minimal()

par(mfrow = c(2, 2))


# histgram of distribution

hist(df$Min.Delay, 
   main = "Delay (in Minutes)", 
   xlab = "Value", 
   ylab = "Frequency", 
   col = "#ADD8E6", 
   freq = FALSE) # `freq = FALSE` scales the histogram for overlaying density
lines(density(df$Min.Delay))

hist(df$Min.Gap, 
   main = "Gap (in Minutes)", 
   xlab = "Value", 
   ylab = "Frequency", 
   col = "#ADD8E6", 
   freq = FALSE) # `freq = FALSE` scales the histogram for overlaying density
lines(density(df$Min.Gap))

```

There are few delays and gaps with time greater than 50 minutes, they might be rare and standalone events, so we drop them.


```{r}
df <- df[df$Min.Delay <= 50 & df$Min.Gap <= 50, ]

```

```{r}
# Gather Min.Delay and Min.Gap into one long-format data frame
delay_data <- data.frame(
  Category = rep(c("Min.Delay", "Min.Gap"), each = nrow(df)),
  Value = c(df$Min.Delay, df$Min.Gap)
)


# Plot
ggplot(delay_data, aes(x = Category, y = Value)) +
  geom_boxplot(fill = "lightgreen") +
  labs(title = "Boxplot of Min.Delay and Min.Gap",
       x = "",
       y = "Minutes") +
  theme_minimal()

```


```{r}
stat.desc(df[, c("Min.Delay", "Min.Gap")])

```
  
  
Checking for correlation: 

```{r}
cr <- cor(df[sapply(df, is.numeric)], method="spearman")
round(cr, 2)

```
Since Delay of one train and the Gap between trains are highly correlated, we can drop one. Here we chose to drop Min.Gap.

```{r}
df <- df[, -5]
summary(df)
```

```{r}
par(mfrow = c(2, 2))

hist(df$Min.Delay, 
   main = "Delay (in Minutes)", 
   xlab = "Value", 
   ylab = "Frequency", 
   col = "#ADD8E6", 
   freq = FALSE) # `freq = FALSE` scales the histogram for overlaying density
lines(density(df$Min.Delay))

```

Since there are a lot of reported incidents that have caused 0 mins delay, it might be a good idea to remove them from our data frame as it skews our data and can create bias in our models later on.

```{r}
#df <- df[df$Min.Delay > 0 & df$Min.Gap > 0, ]

par(mfrow = c(2, 2))

hist(df$Min.Delay, 
   main = "Delay (in Minutes)", 
   xlab = "Value", 
   ylab = "Frequency", 
   col = "#ADD8E6", 
   freq = FALSE) # `freq = FALSE` scales the histogram for overlaying density
lines(density(df$Min.Delay))


```


Now we can see that in *Bound*, the null values are just 125 and in *Vehicle* the null values are 130. This is not a significant number so we drop them as well.

```{r}
df$Bound <- as.character(df$Bound)
df$Vehicle <- as.character(df$Vehicle)

# Remove rows with invalid Bound or Vehicle
df <- df[df$Bound != "" & df$Vehicle != "0", ]

df$Bound <- as.factor(df$Bound)
df$Vehicle <- as.factor(df$Vehicle)

```


## Preparing for Cassification

We create a new column with all values greater than 5 mins as True and less than 5 mins as False. After that we drop the column with time of delays. 

```{r}
df$DelayOver5 <- df$Min.Delay > 5

df <- df[, -5]



```
Since, we will just look at WeekEnd or WeekDay we also won't need Date as we are not considering public holidays or events.

```{r}
### Droping Date column
df <- df[, -1]
```




## What are the most frequent incidents (Code)?

```{r}
code_freq <- table(df$Code)

top5 <- sort(code_freq, decreasing = TRUE)[1:5]

top5_df <- data.frame(Code = names(top5), Frequency = as.numeric(top5))

ggplot(top5_df, aes(x = reorder(Code, Frequency), y = Frequency)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 5 Delay Causes",
       x = "Delay Code",
       y = "Number of Delays") +
  theme_minimal()


```
```{r}

code_legend <- read.csv("Code Descriptions.csv", stringsAsFactors = FALSE)
top5_description <- merge(top5_df, code_legend, by = "Code")
top5_description

```

The most common causes of delays are disorderly patron, injured/ill customer in station or some other security concern.


## Which stations have the most delays?

```{r}
station_counts <- sort(table(df$Station), decreasing = TRUE)
top_stations <- data.frame(Station = names(station_counts), Frequency = as.numeric(station_counts))

ggplot(head(top_stations, 10), aes(x = reorder(Station, Frequency), y = Frequency)) +
  geom_bar(stat = "identity", fill = "tomato") +
  coord_flip() +
  labs(title = "Top 10 Delayed Stations",
       x = "Station",
       y = "Number of Delays") +
  theme_minimal()

```



# Classification: Predict whether delay duration will be greater than 5 mins 

```{r}
# 80% train and 20 % test
sr<- 0.80

# Find number of rows of data
n.row <- nrow(df)
n.row
```
#### Choosing rows for the training sample
  
```{r}

set.seed(8848) 
training.rows <- sample(1:n.row, sr*n.row, replace=FALSE)

#Assign to the training sample

train <- subset(df[training.rows,])

#Assign the balance to the test sample

test <- subset(df[-c(training.rows),])

summary(train)
summary(test)
```
#### Looking at correlation

```{r}
# Keep only variables with manageable cardinality
#df_clean <- df[, c("Line", "TimeRange", "Bound", "Code", "WeekGroup", "DelayOver5")]

#hc <- hetcor(df)

#round(hc$correlations, 2)
```

```{r}
day_plot <- table(df$DelayOver5, df$WeekGroup)
time_plot <- table(df$DelayOver5, df$TimeRange)
bound_plot <- table(df$DelayOver5, df$Bound)
line_plot <- table(df$DelayOver5, df$Line)
station_plot <- table(df$DelayOver5, df$Station)
code_plot <- table(df$DelayOver5, df$Code)

```

<br>
```{r}
par(mfrow=c(2,2))
barplot((day_plot),beside=T)
barplot((time_plot),beside=T)
barplot((bound_plot),beside=T)
barplot((line_plot),beside=T)

par(mfrow=c(1,1))

```

<br>


<br>
```{r}

barplot((station_plot),beside=T)
barplot((code_plot),beside=T)
```

<br>


# Model Development

## 1 Full Model using all variables (dependent variable is DelayOver5)
```{r, include=FALSE}
start <- Sys.time()

glm.all <- glm(DelayOver5 ~ .,
               data = train,
               family="binomial")

end <- Sys.time()

glm_all_summary <- summary(glm.all)
glm_all_summary

```

```{r, echo=FALSE}

# Model summary

fisher_iterations_all <- glm_all_summary$iter

aic_value_all <- glm_all_summary$aic

residual_deviance_all <- glm_all_summary$deviance

p_values_all <- glm_all_summary$coefficients[, "Pr(>|z|)"]

intercept_p_value_all <- p_values_all["(Intercept)"]
intercept_status_all <- ifelse(intercept_p_value_all >= 0.05, "Fail", "Pass")

other_p_values_all <- p_values_all[names(p_values_all) != "(Intercept)"]

passed_count_all <- sum(other_p_values_all < 0.05)
total_count_all <- length(other_p_values_all)

cov_stat_all <- ifelse(fisher_iterations_all <= 1.5*ncol(train),"It converged","It does not converge")
cat("Number of Fisher Scoring Iterations:", fisher_iterations_all, ": **",cov_stat_all,"**\n")
print("### using passing standard of: number of iterations <= (1.5 * number of variables); as discussed in class ###")
cat("AIC: **", aic_value_all, "**\n")
cat("Residual Deviance: **", residual_deviance_all, "**\n")
cat("Intercept Pr(>|z|): **", intercept_status_all, "**\n")
cat("Other variables Pr(>|z|): **", passed_count_all, "out of", total_count_all,"Passed **\n\n" )

time_all <- end - start
time_all
```
<br>


### Looking for signigficant data points

```{r}
# All variables
plot(glm.all, which=4, id.n=6)

```

```{r}
cat("Training\n")
pred_all <- predict(glm.all, newdata=train, type="response")
Class_all <- ifelse(pred_all < 0.6,"Over5","Below5")
CF_all <- table(train$DelayOver5, Class_all)
Acc_all <- (CF_all[1,1]+CF_all[2,2])/sum(CF_all)

CF_all
cat("Accuracy: **", round(Acc_all,4), "**\n\n")

cat("Testing\n")
# pred_allT <- predict(glm.all, newdata=test, type="response")
# Class_allT <- ifelse(pred_allT < 0.6,"Over5","Below5")
# CF_allT <- table(test$DelayOver5, Class_allT)
# Acc_allT <- (CF_allT[1,1]+CF_allT[2,2])/sum(CF_allT)
# 
# CF_allT
# cat("Accuracy: **", round(Acc_allT,4), "**\n\n")

```


## 2 Stepwise Model 

It will look at both (forward and backward) process and give us the best one.

```{r, include=FALSE}
start <- Sys.time()

glm.mod = glm(DelayOver5 ~ .    ,
              family="binomial", data=train, na.action=na.omit)
stp.glm <- step(glm.mod, direction="both", trace=TRUE)

end <- Sys.time()

glm_stp_summary <- summary(stp.glm)
glm_stp_summary

```

<br>  


### Model Interpretations and Comments
```{r, echo=FALSE}

# Model summary

fisher_iterations_stp <- glm_stp_summary$iter

aic_value_stp <- glm_stp_summary$aic

residual_deviance_stp <- glm_stp_summary$deviance

p_values_stp <- glm_stp_summary$coefficients[, "Pr(>|z|)"]

intercept_p_value_stp <- p_values_stp["(Intercept)"]
intercept_status_stp <- ifelse(intercept_p_value_stp >= 0.05, "Fail", "Pass")

other_p_values_stp <- p_values_stp[names(p_values_stp) != "(Intercept)"]

passed_count_stp <- sum(other_p_values_stp < 0.05)
total_count_stp <- length(other_p_values_stp)

cov_stat_stp <- ifelse(fisher_iterations_stp <= 1.5*ncol(train),"It converged","It does not converge")
cat("Number of Fisher Scoring Iterations:", fisher_iterations_stp, ": **",cov_stat_stp,"**\n")
print("### using passing standard of: number of iterations <= (1.5 * number of variables); as discussed in class ###")
cat("AIC: **", aic_value_stp, "**\n")
cat("Residual Deviance: **", residual_deviance_stp, "**\n")
cat("Intercept Pr(>|z|): **", intercept_status_stp, "**\n")
cat("Other variables Pr(>|z|): **", passed_count_stp, "out of", total_count_stp,"Passed **\n\n" )

time_glm <- end - start
time_glm
```
### Looking for signigficant data points

```{r}
# Stepwise Model
plot(glm.mod, which=4, id.n=6)
```
There are highly influential datapoints (i.e. beyond Cook's distance)



<br>


## Comparing the two models

```{r}

###################################
## Confusion Matrix and Accuracy ##

cat("Training\n")
pred_glm <- predict(stp.glm, newdata=train, type="response")
Class_glm <- ifelse(pred_glm < 0.6,"Over5","Below5")
CF_glm <- table(train$DelayOver5, Class_glm)
Acc_glm <- (CF_glm[1,1]+CF_glm[2,2])/sum(CF_glm)

CF_glm
cat("Accuracy: **", round(Acc_glm,4), "**\n\n")

cat("Testing\n")
# pred_glmT <- predict(stp.glm, newdata=test, type="response")
# Class_glmT <- ifelse(pred_glmT < 0.6,"Over5","Below5")
# CF_glmT <- table(test$DelayOver5, Class_glmT)
# Acc_glmT <- (CF_glmT[1,1]+CF_glmT[2,2])/sum(CF_glmT)
# 
# CF_glmT
# cat("Accuracy: **", round(Acc_glmT,4), "**\n\n")

# Error in model.frame.default(Terms, newdata, na.action = na.action, xlev = object$xlevels) :
# factor Code has new levels PUEWZ, TUDOE


```
####### Error while splitting data


<br>

# 2 Naive-Bayes Classification
```{r, warning=FALSE}

train$DelayOver5 <- factor(ifelse(train$DelayOver5 == TRUE, "Yes", "No"))

start <- Sys.time()

NB.mod <- NaiveBayes(DelayOver5 ~ ., 
                     data = train, na.action=na.omit)

end <- Sys.time()
time_NB <- end - start

###################################
## Confusion Matrix and Accuracy ##

cat("Training:\n")
pred_NB <- predict(NB.mod, newdata=train)
CF_NB <- table(Actual=train$DelayOver5, Predicted=pred_NB$class)
Acc_NB <- (CF_NB[1,1]+CF_NB[2,2])/sum(CF_NB)

CF_NB
cat("Accuracy: **", round(Acc_NB,4), "**\n")

cat("\nTesting:\n")
pred_NBt <- predict(NB.mod, newdata=test)
CF_NBt <- table(Actual=test$DelayOver5, Predicted=pred_NBt$class)
Acc_NBt <- (CF_NBt[1,1]+CF_NBt[2,2])/sum(CF_NBt)

CF_NBt
cat("Accuracy: **", round(Acc_NBt,4), "**\n\n")

time_NB

```

<br>
 
# 3 Recursive Partitioning Analysis
```{r}
start <- Sys.time()
# 
RP.mod <- ctree(DelayOver5 ~ ., data=train)
# 
end <- Sys.time()
# 
plot(RP.mod, gp=gpar(fontsize=7))
RP.mod
# 
# ###################################
# ## Confusion Matrix and Accuracy ##
# 
# cat("Training:\n")
# pred.RP <- predict(RP.mod, newdata=train)
# CF_RP <- table(Actual=train$DelayOver5, Predicted=pred.RP)
# CF_RP
# Acc_RP <- (CF_RP[1,1]+CF_RP[2,2])/sum(CF_RP)
# cat("Accuracy: **", round(Acc_RP,4), "**\n")
# 
# cat("\nTesting:\n")
# pred.RPt <- predict(RP.mod, newdata=test)
# CF_RPt <- table(Actual=test$DelayOver5, Predicted=pred.RPt)
# CF_RPt
# Acc_RPt <- (CF_RPt[1,1]+CF_RPt[2,2])/sum(CF_RPt)
# cat("Accuracy: **", round(Acc_RPt,4), "**\n\n")
# 
# time_RP <- end - start
# time_RP

```

<br>
 

# 4 Neural Network Fitting
```{r}
# start <- Sys.time()
# 
# nn.mod <- nnet(DelayOver5 ~ .,
#                data=train,
#                size=4,
#                rang=0.0000001, #variance of the starting random weights
#                maxit=1200,
#                trace=FALSE)
# 
# end <- Sys.time()
# 
# nn.mod
# 
# ###################################
# ## Confusion Matrix and Accuracy ##
# 
# cat("\nTraining:\n")
# pred.nn <- predict(nn.mod, newdata = train, type="class")
# CF_NN <- table(Actual = train$DelayOver5, Predicted=pred.nn)
# CF_NN
# Acc_NN <- (CF_NN[1,1]+CF_NN[2,2])/sum(CF_NN)
# cat("Accuracy: **", round(Acc_NN,4), "**\n")
# 
# cat("\nTesting:\n")
# pred.nnt <- predict(nn.mod, newdata = test, type="class")
# CF_NNt <- table(Actual = test$DelayOver5, Predicted=pred.nnt)
# CF_NNt
# Acc_NNt <- (CF_NNt[1,1]+CF_NNt[2,2])/sum(CF_NNt)
# cat("Accuracy: **", round(Acc_NNt,4), "**\n\n")
# 
# time_NN <- end - start
# time_NN

```
  <br>



